---
title: "DESeq Analysis"
format: html
editor: source
---

First we will load the necessary libraries.


```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(magrittr)
```


The data has been extracted from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226134 from the GSE226134_CK_10__norm.xlsx file. 
```{r load data, cache=TRUE}
data <- read.csv("data/input/JoinedWide.csv")

property <- as.data.frame(cbind(data$SegmentDisplayName, data$METASTATIC))
names(property)=c("SegmentDisplayName","METASTATIC")
normCountData <- data[,59:ncol(data)]
row.names(normCountData) <- data$SegmentDisplayName
```




Typical log-2 fold change scores are around 1.5-2. Here we need to drastically increase in order to be more specific for our results.

```{r}
# Load the limma library
library(limma)
library(edgeR)
design <- model.matrix(~property$METASTATIC)

y <- DGEList(counts=t(normCountData),group=property$METASTATIC)
keep <- filterByExpr(y)
y <- y[keep,,keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
v <- voom(y,design,plot=TRUE)

fit <- lmFit(v,design)

contrast <- c(-1,1)
fit2 <- contrasts.fit(fit,contrast)
fit2 <- eBayes(fit2)


de_genes <- topTable(fit2,coef=1,number=Inf,adjust.method="fdr",sort.by="B",lfc=log2(250))

threshold <- 0.05/9223
sig_genes <- de_genes %>% filter(P.Value < threshold)
nrow(sig_genes) #this is saying that all the genes are differentially expressed
```


To check our findings, we have use a t-test

```{r t tests}
#reference from lab 7 
boolM = property$METASTATIC=="True"
ttests = apply(normCountData,2,function(x){t.test(x[boolM],x[!boolM])$p.value})

ttests <- ttests %>% 
  as.data.frame() %>%
  set_colnames("pval") %>% 
  rownames_to_column("gene") %>%
  arrange(pval)

pval_threshold <- 0.05/nrow(ttests) #bonferroni corrections

DEGenes_ttest <- ttests %>% filter(pval < pval_threshold)
nrow(DEGenes_ttest) #so there are 11 differentially expressed genes
```

Similar to a t-test, we can use the non-parametric Wilcoxon rank sum test to see differentially expressed genes.

```{r wilcoxon tests, warning=FALSE}
# Define a custom function to perform wilcoxon rank sum test and return p-value
#reference from lab 7 
boolM = property$METASTATIC=="True"
wilcoxon_tests = apply(normCountData,2,function(x){wilcox.test(x[boolM],x[!boolM], alternative="two.sided")$p.value})

wilcoxon_tests <- wilcoxon_tests %>% 
  as.data.frame() %>%
  set_colnames("pval") %>% 
  rownames_to_column("gene") %>%
  arrange(pval)

pval_threshold <- 0.05/nrow(wilcoxon_tests) #bonferroni corrections

DEGenes_wilcoxon_tests <- wilcoxon_tests %>% filter(pval < pval_threshold)
nrow(DEGenes_wilcoxon_tests) #so there are 2 differentially expressed genes

```


```{r compare results}
DEGenes_wilcoxon_tests$gene %in% DEGenes_ttest$gene
DEGenes_ttest$gene %in% row.names(sig_genes)
DEGenes_wilcoxon_tests$gene %in% row.names(sig_genes)
```

